---
title: "R Cheatsheet BIO311"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#install.packages("tinytex")
#library(tinytex)
#install.packages("mosaic") 
#include=FALSE means that code and output wil be hidden from final document, but code will still run. (eval=FALSE will make the code not run).
library(mosaic)
```


# READING IN DATA

### read.csv()
The `read.csv` function takes a csv file from your computer and uploads it into R
```{r, echo=FALSE}
bumpusdata<-read.csv("BIO311_data_bumpus.csv")
males<-subset(bumpusdata, bumpusdata$sex=="male")
females<-subset(bumpusdata, bumpusdata$sex=="female")
```

```{r, eval=FALSE, error=FALSE}
bumpusdata<-read.csv("BIO311_data_bumpus.csv")
```

### rm(list=ls())
`rm(list=ls())` clears R's memory, clearing anything you have named, or any other code you have run.

### getwd()
The `getwd()` function will tell you what working directory is running in R. Great for trouble-shooting

### setwd()
On a Mac, I prefer to use command+D to set the working directory, but this code works if your computer is being stubborn or you need to change your working directory for a specific section

```{r, eval=FALSE, error=FALSE}
setwd("~/Desktop/Biostats/Files")
```

# MATRIX MANIPULATION
### matrix(c( )nrow= )
The `matrix(c( ), nrow= )` function is used to create a matrix within the c() function list the numbers down the columns of the matrix
```{r}
a<-matrix(c(1,2,3,4,5,6,7,8,9,10,11,12,13,14), nrow=2)
a
```

### t()
The `t()` function transposes the matrix, so column 1 -> row 1, column 2 -> row 2, column 3 -> row 3, etc. in the transposed matrix.

Before transpose
```{r}
b<-matrix(c(2,4,6,8,10,12,14,16,18,20,22,24,26,28), nrow=2)
b
```

After transpose
```{r}
tb<-t(b) 
tb
```

### cbind()
The function `cbind()` allows you to add two matrices together by the columns (pushing matrices together side-by-side)
```{r}
cbind(a,b)
```

### rbind()
The function `rbind()` allows you to add two matrices together by the rows. (stacking matrices on top of each other)
```{r}
rbind(a,b)
```

### is.matrix
`is.matrix` function "asks" R if it is understanding an object/list as a matrix. 

* An output of TRUE means that the object is being read as a matrix
* An output of FALSE means that the object is being NOT read as a matrix
```{r}
is.matrix(a)
```

### as.matrix
`as.matrix` function "commands" R to understanding an object/list as a matrix, even if it is not supposed to be a matrix.

Originally, "not_M" is not being read as a matrix
```{r}
not_M<-c(1,2,3,4,5,6,7,8,9,1,0,11,12)
is.matrix(not_M)
```

...but after using as.matrix, not_M is being read as a matrix.
```{r}
new_not_M<-as.matrix(not_M)
is.matrix(new_not_M)
```



### %*%    

The `%*%` function multiplies two matrices together
```{r}
a%*%tb
```

# MANIPULATING DATA

### $
The `$` isn't a function, but it is very useful because it allows you to isolate a column from a matrix.
```{r}
bumpusdata$sex
```

```{r}
bumpusdata$wgt
```

### subset()
The `subset()` function allows you to create a more specific matrix from a larger one, based on one variable. For example, the "males" subset below, sifts through "bumpusdata" and picks out all the individuals whose sex is male.
```{r}
males<-subset(bumpusdata, bumpusdata$sex=="male")
```

### head()
The `head()` function lets you see the title of each column and the first 6 rows of the datasheet
```{r}
head(males)
```

### length()
The `length()` function finds the length of a vector. Uou have to specify a column, because if you include an entire matrix (with rows **and** columns) within the argument R will only count the number of columns.

* You have to pair males, with survival (or any other column in bumpusdata), otherwise R will count horizontally (the number of coumns), instead of counting the number of rows.
    + There are 87 males
```{r}
length(males$surv)
```

### summary()
The `summary()`function allows you to view descritive statistics from your dataset. 
```{r}
summary(males$wgt)
```

### c()
The `c()` function allows you to link together a string of individual (numbers, colors, labels, etc.) to create a vector

* When creating a vector with words, you need to put each word into quotations.
```{r}
age<-c(11,14,12,9,11,16,8,10,13,6,5)
names<-c("Sally","Mark","Jeff", "Bill", "Sarah", "Sam", "Karen", "Krista", "Amy", "Mathew", "Caroline")
```

```{r}
age
names
```

### paste()
The `paste()` function takes two identifying factors of an individual, and combines them together into one identifying factor.
```{r}
class<-paste(age,names)
class
```

### rep()
The `rep()` function repeats a specific number or word as a vector a specific number of times.

* This function repeats the word "RED", "BLUE", and "YELLOW" 25 times each.
```{r}
c(rep("RED",25), rep("BLUE",25), rep("YELLOW", 25))
```

### which()
The `which()` function will tell you the position of numbers in a vector that satisfy a certain condition

* Say we have 9 children in a single file line. The first child is 12 years old, the second is 11 years old, etc. Each child has a specific spot in relation to the other children. Using the which() function I can find the position of children over the age of 10. There are 6 children that are over the age of 10

`which(vector,condition,value)`
```{r}
children<-c(12,11,13,18,9,11,7,5,15)
which(children>=10)
```
By combining the which() and length() functions I can find the total number of children that are   over (or under using <=) the age of 10.
    
```{r}
length(which(children>=10))
```

### set.seed() and rnorm()
`set.seed` is R's random number generator, this is useful for creating simulations or random objects that can be **reproduced**
`rnorm()` generates a vector of normally distributed random numbers. 

* by combining set.seed() and rnorm(), you can create unique set of normally distrubued numbers.  
    * The semicolon allows you to perform two "lines of code" on a single line. It is important not to overuse this function, it is much easier to stay organized in R when using seperate lines for seperate functions.
    
Below is simulated data showing the ages of residents in two retirement homes
```{r}
set.seed(12345); retire1_data<-rnorm(n=1000,mean=90, sd=3.3)
set.seed(54321); retire2_data<-rnorm(n=1000,mean=86, sd=5.1)
```

### tapply()
Using the `tapply()` function requires continuous data,the categories and the calculation you want to do. tapply(continuous, categories, function)

Using the simulated data for ages of residents in two retirement homes, we can find the mean of the two goups using the tapply function  

* It is important to notice how retirement_ages and groups are the same length (meaning they both have 1000 individuals)
```{r}
retirement_ages<-c(retire1_data, retire2_data)
groups<-c(rep("retire1",1000), rep("retire2",1000))
tapply(retirement_ages, groups, mean)
```

### sample()
the `sample()` function can take a sample from a data set. In this case we have a bowl with 15 red marbles and 20 blue marbles  

sample(dataset, number of individuals in each sample)
```{r}
bowl<-c(rep("RED", 15), rep("BLUE", 20))
sample(bowl, 5)
```


### loops
Here is the general "formula" for loops

**output<-NULL**  
**n<- #number of times the loop runs**  
**for(i in 1:n){**  
  **#tasks...enter code you want to run repeatedly**  
  **output<-c(output, task result) #this line puts the outputs in a list**  
**}**

For example, say I have a bowl of 500 seeds with an average mass of 50 mg. What is the probability that if I sample 2 seeds from the bowl, the average will be greater than 50?
```{r}
set.seed(2); seeds<-rnorm(n=500,mean=50,sd=10)
safe<-NULL
n<-100
for(i in 1:n){
  two_seeds<-sample(seeds, 2)
  avg.seed.wgt<-mean(two_seeds)
  safe<-c(safe, avg.seed.wgt)
}

safe

#Outside the loop, you can figure out the probability of pair of seeds being over 50 milligrams

sum(safe>=50)/n
```
From our answer we can see that 53% of the seeds sampled had a mean greater than the population mean

### table()
The `table()` function builds a contingency table of the counts for each combination of categorical factors
```{r}
bumpustable<-table(bumpusdata$sex, bumpusdata$surv)
bumpustable
```

# TESTING ASSUMPTIONS
## Normality
### qqnorm()
The `qqrnorm()` is a normality visualization that creates a scatter plot of your data vector against expected values of data if they  were *normally distributed* about the mean. If the data is normally distributed, the data will be in a straight linear line.

For example, the male bird weight from the bumpus data. (the "males" subset was made previously)

```{r}
qqnorm(males$wgt)
```

### shapiro.test()
The Shapiro-Wilk tests if a data vector is normally distributed. If the test is significant, then the data do NOT come from a normally distributed population. Remember this by thinking that the data is "significantly different from a normal population"

For example, the male bird weight from the bumpus data.
```{r}
shapiro.test(males$wgt)
```
The male bird weights from the bumpus data is NOT normally distribued because we got a significant shapiro-wilk test.

## Equal variance
### var.test()
When the data are normal, the F-test compares the variances of 2 groups. If the test is significant, the variances of the two groups are NOT equal. Remember this by thinking that the data is "significantly different from having equal variance." The F-test can only compare two groups.

For example, the male and female bird weight from the bumpus data.
```{r}
var.test(males$wgt,females$wgt)
```
The test is NOT signfificant, so the two data sets have equal variance.

### leveneTest()
When the data are NOT normal, the levene test compares the variances between 2+ groups. If the test is significant, the variances of the groups are NOT equal

To use the levene test, you have to first, download the package "car".
```{r, message=FALSE}
# install.packages('car')
library(car)
```
For example, the male weight (y) and male survival (x) from the bumpus data.
```{r}
leveneTest(males$wgt~males$surv)
```
The result from the levene test is not significant so the groups (male alive and male dead) have equal variances.


# INFERENTIAL STATISTICS

### dbinom()
`dbinom(x, size, prob)`
**x= a vector of numbers, size=number of trials, prob= probability of success for each trial**  
Using a bionomial probability distribution, you can use the dbinom() function to calculate the probability of exactly one outcome under that null distribution.

* Returning to our simulated retirement home data, I want to know if I knock on 20 doors in the retirement home, what are the chances that exactly 5 95-year-olds will answer their door (The probability of success i.e. finding a 95-year-old is 0.055)? 
```{r}
dbinom(5,20, 0.055)
```
NOTE: this is NOT a p=value, but a probability of a very specific outcome (finding exactly 20 95-year-olds out of the 1000 residents)

### binom.test()
`binom.test(number of sucesses, number of trials)`   
p=expected/hypothesized probability of success, alternative="two.sided" "greater" or "less")**

You can use binom.test() function to calculate the probability of success under a binomial probability distribution. Where dbinom() only calculates the probability of one specific outcome/sucess, binom.test() calculates a range of outcomes that are all considered a success.

* Again returning to the retirement home data, if I knock on 20 doors in the retirement home, what are the chances that I will find five people who are 95-years-old OR OLDER will answer their door (The probability of success i.e. finding a 95-year-old is 0.061)? 
```{r}
binom.test(5, 20, p=0.061, alternative="greater")
```

### chisq.test()
`chisq.test(x, p=)` is a chi-squared goodness-of-fit test (where x is data and p is null probability)

* An example of this the different amounts of each color of skittle in the bag. (blue=103, red=61, green=55, orange=64, yellow=77)
```{r}
chisq.test(c(103,61,55,64,77), p=c(0.2,0.2,0.2,0.2,0.2))
```
chisq.test(x) is a chi-squared test-of-association (where x is the data)

* Looking back to the bumpus data, did sex of the bird influence if it lived or died? We can answer this using a chi-squared test-of-association. (The expected probability is derived from the observed data, so thats why its not in the function)
```{r}
bumpustable<-table(bumpusdata$sex, bumpusdata$surv)
chisq.test(bumpustable)
```

###g.test()
Similar to a chi-squared test, but... useful for complicated experimental design
ex: null hypothesis is a particular association
Does not function well with small sample sizes
Use the same degrees of freedom as with chi-squared
    *You can set your expected frequencies in a way you can't with chi-squared
G=2(Obs*ln(O-E))


### t.test()
Unsurprisingly, the function, t.test is used to perform a t-test on a dataset. A t-test can determine if there is a significant difference between the means of two goups (with continuous data).  

There are 3 different types of t-tests:     

***One-sample t-test** Also called the "Goodness of Fit t-test." How the population mean compares to some hypothesized value. Example: Imagine we want to compare this years flow of the Nile river compared to the average flow over the last 20 years (800.53)  
  * (The dataset Nile is pre-loaded into R)
    ```{r}
Nile_past<-800.53
t.test(Nile, mu=800.53)
```
Based on this one-sample t.test, there is a significant difference between the mean flow of the Nile river this year, to the mean over the last 20 years.

**Two-sample t-test** How the means of two groups are different. For example, if we want to see if there is a significant difference in the weight's of crickets after feeding them either food-x or food-y  

* (The dataset randu is pre-loaded into R)
    
```{r}
t.test(randu$x,randu$y)
```
Based on this two-sample t.test, there is a significant difference between the weights of the crickets under the two different food groups.  

**Paired t-test** Determines whether the mean of two different groups is 0 (null hypothesis) or not (alternative). For example, the number of pimples before and after a cream is applied from a dermatologiest.

```{r}
set.seed(4); before_cream<-rnorm(10, mean=10, sd=6 )
set.seed(4); after_cream<-rnorm(10, mean=9, sd=7)
t.test(before_cream, after_cream, paired=TRUE)
```
Based on this paired t.test, there is NO significant difference between the number of pimples before and after the cream. Maybe time to find a new dermatologist...

### summary(aov())
The `summary(aov())` function above us used to conduct and ANOVA (Analysis of Variance) which is a test that uses the difference in variation to see if the mean between groups are different  

* Unlike the two-sample t.test, you can have more than two groups

* For example, determing if plants have different CO2 intake based on their type (Quebec or Mississippi) AND treatment (chilled or nonchilled)  

* **IMPORTANT- ANOVA's do NOT tell you WHICH of the groups are different, only that a difference exists between at least two of the groups.**  

    + CO2 is a preloaded dataset in R
    
```{r}
PLANT<-paste(CO2$Type, CO2$Treatment)
summary(aov(CO2$uptake~PLANT))
```

### TukeyHSD(aov()) 
A **Tukey test** must be used to identify the groups that have a significant difference. The output is a list of p-values of the differences between each of the pairings.
```{r}
TukeyHSD(aov(CO2$uptake~PLANT))
```
Focusing on the P values in the reults you can see there is a significant differene between all of the groups except between Quebec nonchilled-Quebec chilled (p=0.4727)


# PLOTTING DATA
### hist()
The `hist()` function creates a histogram of the weight of male birds.
```{r}
hist(males$wgt)
```

### boxplot()
Creates a boxplot of bird weight and "male alive" "male dead" "female alive" "female dead"  

* The generic formula for a boxplot is `boxplot(y~x)`
    + y= continuous
    + x= categorical
    
Notice the paste function!
```{r}
y<-bumpusdata$wgt
x<-paste(bumpusdata$surv,bumpusdata$sex)
boxplot(y~x)
```

### barplot()
Plot with a categorical and continuous variables
```{r}
barplot(c(223,251,317,636,766,607,607), main="Pages in Harry Potter books", ylab="Pages", xlab="Books", names=c("PS","CS","PA","GF","OP","HBP","DH"))
```



# PLOT MODIFIERS
### main=
The `main=` modifier titles the plot
```{r}
boxplot(y~x, main="Bird weight by sex and survival")
```

### ylab=
The `ylab` modifier labels the y-axis
```{r}
boxplot(y~x, main="Bird weight by sex and survival", ylab="Bird weight")
```

### xlab=
The `xlab=` modifier labels the x-axis
```{r}
boxplot(y~x, main="Bird weight by sex and survival", ylab="Bird weight", xlab="Groups")
```

### xlim=
The `xlim=` modifier allows you to change the limits of the x-axis
```{r}
hist(males$wgt, xlim=c(20,35))
```

### ylim=
The `ylim=` modifier allows you to change the limits of the x-axis. Here I increased the y-axis to 40
```{r}
hist(males$wgt, ylim=c(0 ,40))
```

### col=
The `col=` modifier allows you to color the plot. (Arguably, the best function within R)

**Color is a tool!** If you are using multiple colors, don't just randomly color in boxes, use the color to easily distinguish between groups. 

* In my plot the females are pink and the males are blue. Additionally, the alive individuals are a lighter shade and the dead individuals are a darker color.
```{r}
boxplot(y~x, main="Bird weight by sex and survival", ylab="Bird weight", xlab="Groups", col=c("palevioletred1","lightskyblue1", "palevioletred4", "lightskyblue4"))
```

### col=rgb()
`col=rgb()` is more specific than the col= modifier Not to be confused with the notorious RBG, col=rgb() allows you to create custom colors by changing the proportion of red, blue and green in that color.

* For example, looking at the bumpus data for male weight, I can change the color and transparancy of the graph

* The values range from 0 to 1, with 1 being the most extreme and 0 being the least extreme
```{r include=FALSE}
females<-subset(bumpusdata, bumpusdata$sex=="female")
```

```{r}
hist(males$wgt, col=rgb(0.1,0.5,0.792,1))
```

### add=TRUE
Using the `add=TRUE` you can combine two histogram code lines into one graph. The add=TRUE is inserted as a modifier into the second hist(). Any modifiers to change graph (ylim, xlim, ylab, xlab, main, etc.) must be added to first histogram function

This is a great way to use the col=rgb() function and change the transparancy of the histograms to see where the histograms overlap.
In this example I am using the bumpus dataset to compare male and female bird weight.
```{r}
hist(females$wgt, col=rgb(0.9, 0.2, 0.32, 0.8), ylim=c(0,28), xlab=("bird weight"), main=("Histogram of male and female bird weight"))
hist(males$wgt, add=TRUE, col=rgb(0.1, 0.5, 0.7, 0.55))
```

### segments()
The `segments()` modifier is NOT a modifier, so it requires a new line of code to work. The function basically works by connecting two points with a straight line.

* segments(X~1~, Y~1~, X~2~, Y~2~)
* You can change the color of segments using either col=() or col=rbg()
* lwd() changes the thickness of the line

While walking long the mall, I notice there is a single male bird, the exact species that Bumpus used. What a lucky find! I take his mass. I am interested to see if the bird mass is different now from when Bumpus did his experiment **The segment line is the mean masse of the male bird I caught.**

```{r}
hist(males$wgt)
segments(27.5,0,27.5,25, col="royalblue3", lwd=5, lty="dotted")
```

### text()
`text()` modifier adds text to any plot
```{r}
hist(males$wgt)
segments(0,10,40,10, col="forestgreen", lwd=5)
text(29,11,"This is a line :)")
```

### legend()
The `legend()` modifier puts a legend on your graph, telling the reader what colors are certain variables
legend(), like segments() needs its own line beneath the histogram

* Continuing with the same graph from above...
```{r}
hist(females$wgt, col=rgb(0.9, 0.2, 0.32, 0.8), ylim=c(0,28), xlab=("bird weight"), main=("Histogram of male and female bird weight"))
hist(males$wgt, add=TRUE, col=rgb(0.1, 0.5, 0.792, 0.4))
legend("topright", c("Females", "Males"), fill=c("deeppink2","royalblue3"))
```

# THAT'S ALL FOLKS!



